Improving the portuguese hyphenation rules

### 2023-10-13 article/article.tex
revision of section 1

\section{Introduction}\label{sec-intro}
Hyphenation in text wrapping was not used for a long time. Words should fit
entirely in a line, or they would be broken in arbitrary places. First no
marker at all was used to sign the wrapping of a word, what could create
confusion and undesirable meaning. Therefore, ortographers advocated for the
use of a sign to indicate this break. Portuguese faced the same gradual
introduction of a hyphenation sign to mark words wrapping across lines. Even
though the usage of a hyphenation sign ($=$) was advocated by orthographers
\cite{gandavo1574}, few documents used such sign until the end of the 18th
century \cite{araujo2015}.

In some cases, hyphenation hinders smooth reading and should be avoided in
child literature, for example. In opposition, large or small spaces between
words also impose difficulty in the reading process, making hyphenation
fundamental when texts use short line lengths. As a line gets shorter, the
number of breaking candidates between words decreases, leading to awkward
spaces between words and among letters. For that matter, automatic hyphenation
plays an important role in good typesetting.

@@

\TeX{} is a typesetting system which carefully deals with these issues,
automatically arranging text on a page to create a good reading experience.
Automatic hyphenation is an important part on this process,
promoting an even-tempered distribution of elements on the page.
%Hyphenation is therefore an important part on this process.
%A good readability depends on an even-tempered distribution of elements
%on the page.
Line height and length, paragraph length, font size and typeface,
letter and word-spacing, are some factors which influence the text legibility
and readability. Space between words should not be too long creating lakes and
rivers in the text, nor too tight, impairing the legibility and readability.

@@

The general rule for Portuguese hyphenation is to split a word into its
syllables. A syllable is made of a mandatory nucleus, filled by a vowel, and
optional peripheral consonants (before or after the nucleus). In some
situations, the syllabic division does not respect the ethnologic constituents.
The usage of the prefixes \emph{bis-} and \emph{in-} are examples of this
circumstance. The correct syllabifications are \emph{bi-sa-vô},
\emph{i-na-ti-vo}, and \emph{i-nobs-tan-te}\footnote{The rule of syllable
division could lead to two possible partitions: \emph{i-nobs-tan-te} and
\emph{in-obs-tan-te}, but the first is preferable.}, where the prefixes are
split into two syllables. But, as pointed previously, it is preferable to keep
morphemes together rather than splitting them apart, therefore we should
favor the hyphenation \emph{bisa-vô} over \emph{bi-savô}, and \emph{ina-tivo}
over \emph{i-nativo}. This last example could also lead to a misunderstanding
since the word \emph{nativo} (\emph{native} in English) emerges from this word
break.

Each language has its own hyphenation rules. They might be grouped into two
groups: those which hyphenation is driven by morphology (etymology) and those
driven by pronunciation. An algorithmic approach might use a logic system to
analyse words and apply the hyphenation rules of a given language. An algorithm
must be written for each language, as hyphenation rules vary extensively.
Although a logic system might be fast and compact, inevitably it will have to
deal with exceptions through hard-coded rules.

The automation of this process might use a dictionary based approach, which
will restrict the hyphenation possibilities to those entries in the dictionary,
an algorithmic base approach, which might be applied to whatever sequence found
in a text, or a mix of both approaches. 
%An algorithmic approach might use a logic system to analyse words and apply
%the hyphenation rules of a given language. 
A rule base model might include the recognition of prefixes,
suffixes, morphemes, or some sequences suitable for the inclusion of a break.
Another approach is the usage of pattern matching. Given a corpus of
hyphenation examples in a language, it is necessary to identify those sequence
of letters that define a point where hyphenation is suitable or not. Patterns
might include prefixes, suffixes, exceptions and special hyphenation rules of a
given language. Such approach might be used for different languages, by
providing hyphenation examples in a target language to extract patterns and
rules.

Only the analysis of immediate surrounding might not be enough to determine a
potential break point. For example, consider \emph{dem-o-crat} and \emph{de-moc-ra-cy},
where immediate surrounding of \emph{e} does not suffice to determine the break
location. Even after chosen patterns that makes hyphenation pretty
straightforward, there might be exceptions. Consider the sequence \emph{tion}. It
seems safe to always place a break before this pattern. However, the word
\emph{cation} is hyphenated as \emph{cat-ion}, making is etymology determinant to the way
this word will be split.

A word with different meanings may be hyphenated differently according to its
meaning. `For instance, the Swedish word form \emph{glassko} has three different
meanings, and can be hyphenated as \emph{glas-sko} (glass shoe),
\emph{glass-ko} (ice cream cow) and in the non-standard way,
\emph{glass-sko} (ice cream shoe).' \cite{nemeth2006}. In Portuguese, the
word \emph{sublinha} might be hiphenated in two different forms:
\emph{su-bli-nha} when representing the inflected form of verb
\emph{sublinhar} (underline as a verb) or \emph{sub-li-nha} when refering
to the line under (underline as a noun).

@@

Machine learning approach was used to hyphenate Norwegian text
\cite{kristensen2001}. It was found that the \TeX{} approach, in general,
performed better then a neural network. Both approaches had similar
performance on finding the correct hyphenation points and avoiding wrong
hyphenation points on a small word list (the neural network performed slightly
better on finding the correct hyphenation points). Nonetheless, \TeX{} approach overcame
the neural network on a large word list, regarding the avoidance of wrong
hyphenations.

The original \TeX{} hyphenation algorithm was introduced by Knuth 1977
\cite{knuth1977}. It was focused on the English language and used three main
rules:
\begin{enumerate*}[label=(\arabic*)]
 \item suffix removal; 
 \item prefix removal; and 
 \item vowel-consonant-consonant-vowel (VCCV) breaking. 
\end{enumerate*}
Tests showed it could find
40\% of the allowed hyphen locations \cite{liang1983}. 
The hyphenation algorithm proposed by Frank M. Liang and adopted in \TeX{}
utilizes competing patterns. The \TeX{}82 algorithm employs five alternating
levels of hyphenating and inhibiting patterns. The program for pattern
generation, known as PATGEN, was created by Liang \cite{liangbreitenlohner1999}
and has been utilized to generate hyphenating patterns for numerous languages
\cite{sojka1995,sojka1995a,sojka2005thesis,sojka2003,scannell2003}. It involves
sweeping a database of hyphenated words in a language to identify both
hyphenating and inhibiting patterns, ultimately creating a list of competing
patterns for that specific language.

@@

>>>>

\section{Introduction}\label{sec-intro}

Hyphenation in text wrapping was not used for a long time. Words should fit
entirely in a line, or they would be broken in arbitrary places.  Initially, no
markers were used to indicate word wrapping, leading to potential confusion and
unintended interpretations. As a result, orthographers advocated for the
introduction of a sign to indicate such breaks. Portuguese faced the same
gradual introduction of a hyphenation sign to mark words wrapping across lines.
Even though the usage of a hyphenation sign ($=$) was advocated by
orthographers \cite{gandavo1574}, few documents used such sign until the end of
the 18th century \cite{araujo2015}.

In some cases, hyphenation hinders smooth reading and should be avoided in
child literature. In opposition, both excessive spacing and insufficient
spacing between words also impose difficulty in the reading process, making
hyphenation fundamental when texts use short line lengths. As a line gets
shorter, the number of breaking candidates between words decreases, leading to
awkward spaces between words and among letters. For that matter, automatic
hyphenation plays an important role in good typesetting.

@@

\TeX{} is a typesetting system which carefully deals with these issues,
automatically arranging text on a page to create a good reading experience.
Automatic hyphenation is an important part on this process,
promoting an even-tempered distribution of elements on the page.
%Hyphenation is therefore an important part on this process.
%A good readability depends on an even-tempered distribution of elements
%on the page.
Line height and line length, paragraph length, font size and typeface,
letter-spacing, and word-spacing are some factors which influence the text legibility
and readability. Space between words should not be too long creating lakes and
rivers in the text, nor too tight, impairing the legibility and readability.

@@

The general rule for Portuguese hyphenation is to split a word into its
syllables. A syllable is made of a mandatory nucleus, filled by a vowel, and
optional peripheral consonants (before or after the nucleus). In some
situations, the syllabic division does not respect the ethnologic constituents.
The usage of the prefixes \emph{bis-} and \emph{in-} are examples of this
circumstance. The correct syllabifications are \emph{bi-sa-vô},
\emph{i-nobs-tan-te}\footnote{The rule of syllable
division could lead to two possible partitions: \emph{i-nobs-tan-te} and
\emph{in-obs-tan-te}, but the first is preferable.}, and \emph{i-na-ti-vo}, 
where the prefixes are split into two syllables. 
But, as pointed out previously, it is preferable to keep
morphemes together rather than splitting them apart. Therefore we should
favor the hyphenation \emph{bisa-vô} over \emph{bi-savô}, \emph{inobs-tante} 
over \emph{i-nobstante}, and \emph{ina-tivo} over \emph{i-nativo}. 
This last example could also lead to a misunderstanding
since the word \emph{nativo} (\emph{native} in English) emerges from this word
break.

Each language has its hyphenation rules, which can be categorized into two
groups: those driven by morphology (etymology) and those driven by
pronunciation.  An algorithmic approach employs a logical system to analyze
words and apply the hyphenation rules of a specific language. Since hyphenation
rules vary significantly between languages, a unique algorithm must be
developed for each one. While a logic-based system can be efficient and
compact, it still needs to address exceptions through hard-coded rules.

%The automation of this process might use a dictionary based approach, which
%will restrict the hyphenation possibilities to those entries in the dictionary,
%an algorithmic base approach, which might be applied to whatever sequence found
%in a text, or a mix of both approaches. 
%%An algorithmic approach might use a logic system to analyse words and apply
%%the hyphenation rules of a given language. 
%A rule base model might include the recognition of prefixes,
%suffixes, morphemes, or some sequences suitable for the inclusion of a break.
%Another approach is the usage of pattern matching. Given a corpus of
%hyphenation examples in a language, it is necessary to identify those sequence
%of letters that define a point where hyphenation is suitable or not. Patterns
%might include prefixes, suffixes, exceptions and special hyphenation rules of a
%given language. Such approach might be used for different languages, by
%providing hyphenation examples in a target language to extract patterns and
%rules.

The automation of this process can employ various approaches, including:
\begin{description}
\item[Dictionary-Based Approach:] This method restricts hyphenation
    possibilities to entries found in the dictionary.
\item[Algorithmic-Based Approach:] This approach can be applied to any sequence
    encountered in a text.
\item[Rule-Based Model:] This model recognizes prefixes, suffixes, morphemes,
    or specific sequences suitable for hyphenation.
\item[Pattern Matching:] By using a corpus of hyphenation examples in a
    language, this approach identifies letter sequences that determine suitable
    hyphenation points. Patterns encompass prefixes, suffixes, exceptions, and
    special hyphenation rules of the language.
\item[Mixed Approach:] This approach combines two or more of the previously
    described methods to enhance hyphenation accuracy and flexibility.
\end{description}

Analyzing only the immediate surroundings may not always suffice to determine a
potential hyphenation point. For instance, consider \emph{de-moc-ra-cy} and \emph{dem-o-crat},
where the immediate surrounding of the letter \emph{e} does not provide a clear
indication of the hyphenation point. 
Even when facing chosen patterns that typically make hyphenation
straightforward, exceptions can arise. Take the sequence \emph{tion}, for example.
It may seem logical to place a break before this pattern, but the word \emph{cation}
is hyphenated as \emph{cat-ion}, highlighting the influence of etymology on the way
a word is split.

A word with multiple meanings may have distinct hyphenations based on its intended meaning.
`For instance, the Swedish word form \emph{glassko} has three different
meanings, and can be hyphenated as \emph{glas-sko} (glass shoe),
\emph{glass-ko} (ice cream cow) and in the non-standard way,
\emph{glass-sko} (ice cream shoe).' \cite{nemeth2006}. In Portuguese, the
word \emph{sublinha} might be hiphenated in two different ways:
\emph{su-bli-nha} when representing the inflected form of the verb
\emph{sublinhar} (to underline) or \emph{sub-li-nha} when refering
to the line under (underline as a noun).

@@

Machine learning techniques were employed to hyphenate Norwegian text
\cite{kristensen2001}. The study revealed that, overall, the \TeX{} approach
outperformed a neural network.  Both methods demonstrated similar performance
in identifying correct hyphenation points and minimizing incorrect hyphenations
when tested on a small word list (with the neural network slightly
outperforming in correctly recognizing hyphenation points).  However, when
tested on a larger word list, the \TeX{} approach proved superior in avoiding
incorrect hyphenations compared to the neural network.

The original \TeX{} hyphenation algorithm, introduced by Knuth in 1977
\cite{knuth1977}, primarily focused on the English language and employed three
main rules:
\begin{enumerate*}[label=(\arabic*)]
 \item suffix removal; 
 \item prefix removal; and 
 \item the Vowel-Consonant-Consonant-Vowel (VCCV) breaking rule\footnote{The 
     VCCV rule in hyphenation patterns, places the syllable boundary between 
     two consecutive consonants when they appear between two vowels. For example, 
     in the word \emph{sudden}, the syllable break occurs between the \emph{d} and 
     the second \emph{d}, making it \emph{sud-den}. 
     The VCCV pattern is a relatively common syllable division pattern in English.
     This rule ensures accurate 
     hyphenation, maintaining word readability and pronunciation when words are 
     split at the end of a line in printed text.}. 
\end{enumerate*}
Tests demonstrated that it could identify 40\% of allowable hyphen locations
\cite{liang1983}.  The \TeX{} hyphenation algorithm later adopted the approach
proposed by Frank M. Liang, which involves competing patterns.  The \TeX{}82
algorithm employs five alternating levels of hyphenating and inhibiting
patterns. The program for pattern generation, known as PATGEN, was created by
Liang \cite{liangbreitenlohner1999} and has been utilized to generate
hyphenating patterns for numerous languages
\cite{sojka1995,sojka1995a,sojka2005thesis,sojka2003,scannell2003}. It involves
sweeping a database of hyphenated words in a language to identify both
hyphenating and inhibiting patterns, ultimately creating a list of competing
patterns for that specific language.

@@

Moreover, as mentioned earlier, certain hyphenations may be preferred for
stylistic reasons, or to avoid ambiguity, or for better reading experience.
Some words might have multiple hyphenations, depending on the intended meaning.
In such cases, \TeX{} may not efficiently address the hyphenation challenge, as
it would necessitate case-by-case handling.
